Variance reduction for MCMC methods via martingale representations

### 2020-02-22 aap/Revision/vr-mcmc-revision-new.tex
update
update part 1

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)  Let $\left(\Omega,\mathcal{F},\mathsf{P}\right)$ be a probability space, $(\xi_p)_{p \in \nset}$ be i.i.d. $d_{\xi}-$dimensional random vectors. We denote by $(\mathcal{G}_p)_{p \in \nset_0}$  the filtration generated by $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For  $x \in \rset^d$ and $y \in \rset^{d_\xi}$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+d_{\xi}}$ to $\rset^d$. Then we can define for $l \le p$

@@

Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain with values in $\rset^d$ of the form \eqref{eq:chain_gen}, starting at $X_0 = x$. We write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively. Let us also define by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $1 \leq l \leq q \leq p$ and $y \in \rset^d$, $\mathsf{E}\left[\left |f(X^y_{l,q})\right|^{2}\right]<\infty$. Note that for $q \leq p$ it holds $\mathcal{F}_2^p \subseteq \mathcal{F}_2^q$.

\begin{thm}\label{prop:29032018a1}
For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_2^p$ and $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}a_{q,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where \((X^x_{p})_{p\geq 0}\) is given in \eqref{eq:ula_new} and for all $y \in \rset^d$
\begin{eqnarray}
\label{eq:coeff_mart}
%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \mathbb{N}
\end{eqnarray}
\end{thm}

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p = q=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
for any $f \in \mathcal{F}_2^p$  with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and $f$ from $\mathcal{F}_2^p$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
For any function $f$ from $\mathcal{F}_2^p$,
the orthonormality and completeness
of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any $y \in \rset^d$, 
$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
\[
f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
\]
$\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$. Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^{X^x_q}_{q,q+1})|X^x_{q}] = \mathsf{E}[f(X^{x}_{q+1})|X^x_{q}]$, it holds in \(L^2(\mathsf{P})\) that
\begin{equation}
\label{eq:1-step-decomp}
f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
\end{equation}
where
\[
a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
\]
%we get by
%conditioning on $\mathcal{G}_{q}$,
%\[
%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
%\]
%where
%\begin{equation*}
%\alpha_{q+1,q+1,k}
%=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
%\end{equation*}
%By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$. Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
%\[
%\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
%=a_{q+1,q+1,k}(X^x_{q})
%\]
%where for $y \in \rset^d$
%\[
%a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
%\]
%We thus arrive at
%\begin{align}
%\label{berm:sig_X}
%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]+\sum_{k\geq1}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1}),
%\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]$
is square integrable and has the form
$g(X^x_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{q+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{q+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{q+1})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
 = \mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
\end{align*}
where for $y \in \rset^d$,
\[
a_{q+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,q+1})\phi_{k}(\xi_{l})\right]
\]
Formulas \eqref{eq:1-step-decomp}
and~\eqref{eq:28082017a1} conclude the induction step for $p = q+1$ and hence the proof.
\end{proof}

\begin{cor}
If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
\begin{eqnarray*}
f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for $y \in \rset^d$
\begin{eqnarray*}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}
\end{cor}

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\end{eqnarray}
where
\begin{equation}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
\end{equation}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain
\[
\PE[A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
\end{equation}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

\sergey{N and N0 were defined here before}
\eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}

>>>>

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)  Let $\left(\Omega,\mathcal{F},\mathsf{P}\right)$ be a probability space, $(\xi_p)_{p \in \nset}$ be i.i.d. $m-$dimensional random vectors. We denote by $(\mathcal{G}_p)_{p \in \nset_0}$  the filtration generated by $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For  $x \in \rset^d, y \in \rset^{m}$ and $k \in \nset$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+m}$ to $\rset^d$. Then we can define for $l \le p$

@@

Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain with values in $\rset^d$ of the form \eqref{eq:chain_gen}, starting at $X_0 = x$. We write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.

\begin{thm}\label{prop:29032018a1}
For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, Borel bounded functions $f: \rset^d \rightarrow \rset$ and $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}a_{q,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where \((X^x_{p})_{p\geq 0}\) is given in \eqref{eq:ula_new} and for all $y \in \rset^d$
\begin{eqnarray}
\label{eq:coeff_mart}
%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \nset
\end{eqnarray}
\end{thm}

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p = q=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
The orthonormality and completeness
of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$, 
$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
\[
f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
\]
Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
\begin{eqnarray*}
\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left. \PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left. \PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
\\
&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
\end{eqnarray*}
Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
\begin{equation}
\label{eq:1-step-decomp}
f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
\end{equation}
where
\[
a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
\]
which is the required statement in the case $q = m+1$ and $j=m$.
Now assume that $j<m$.
Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$. Note that $\mathsf{P}$-a.s. it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction. Hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
 = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
\end{align*}
where for $y \in \rset^d$,
\[
a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
\]
Formulas \eqref{eq:1-step-decomp}
and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
\end{proof}

\begin{remark}\label{rem:general_moments} Denote by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $n \leq p$, any indices $1 \leq i_1 < \ldots < i_n \leq p$ and $y \in \rset^d$,
\begin{equation}
\label{def:f_2_p}
\mathsf{E}\left[\left |f(\Phi_{i_n}(\cdot,\xi_{i_n})\circ\Phi_{i_{n-1}}(\cdot,\xi_{i_{n-1}})\circ\dots\circ\Phi_{i_{1}}(y,\xi_{i_{1}}))\right|^{2}\right]<\infty
\end{equation}
Then the statement of Theorem~\ref{prop:29032018a1} remains valid for $f \in \mathcal{F}_2^p$. 
\end{remark}

If  all the functions $\Phi_l,$ $l\geq 1,$ in \eqref{eq:ula_new} are equal, then the condition \eqref{def:f_2_p} reduces to $\PE \left[ f^2(X_q^y)\right] < \infty$ for all $q \leq p$ and $y \in \rset^d$. Let us denote this class of functions by $\mathcal{F}_{2,\text{hom}}^p$.

\begin{cor}
Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
\begin{eqnarray*}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for $y \in \rset^d$
\begin{eqnarray*}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}
\end{cor}

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\end{eqnarray}
where
\begin{equation}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
\end{equation}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
\end{equation}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that
\begin{eqnarray*}
|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.
\end{equation*}
\end{description}
First we analyse the biases of the estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
\paragraph{Squared bias:} Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[M_{K,n}^N(f)\right]=0,
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
%Notice that this remains true also when we substitute
%the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
%with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$

Denote by $\pi($ denotes the probability measure on $\mathbb R^d$
with density $\pi$ of~\eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one-step of the ULA algorithm by
\[
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
\]
For a bounded Borel function~$f$,
we can estimate the conditional bias
similarly to~\cite[Section~4]{durmus:moulines:2017}:
\begin{equation}
\label{eq:06042018a2}
\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\le
\frac{\mathrm{osc}(f)^2}{n}\sum_{p=N+1}^{N+n}
\tvnorm{\delta_x \kerULA_\gamma^{p}-\pi(\cdot)}^2,
\end{equation}
where
$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
$\tvnorm{\mu-\nu}$ denotes the total variation distance
between probability measures $\mu$ and $\nu$, that is,
\[
\tvnorm{\mu-\nu}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)| \eqsp.
\]
Different specific upper bounds for the squared bias
can be deduced from~\eqref{eq:06042018a2}
using results of Section~3 in~\cite{durmus:moulines:2017}
on bounds in the total variation distance. It is known  that under assumptions {\bf (H1)} and {\bf (H2)},   the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Moreover, as shown in Proposition, \tcr{(Appendix)}
\eric{put a precise reference here}
 \begin{equation}
\label{eq:bias}
\tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \tvnorm{\pi-\pi_\gamma} \leq C_2 \sqrt{\gamma}
\end{equation}
for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and  constants $C_1,$ $C_2>0$ independent of $\gamma$  and $n.$
\paragraph{Variance:} An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
under {\bf (H1)} and {\bf (H2)} is derived in Section.  In particular, for any bounded function $f,$
\begin{equation}
\label{eq:var-mc}
\PVar[\pi_n^N(f)]\lesssim (n\gamma)^{-1} \eqsp,
\end{equation}
where $\lesssim$ stands for inequality up to a constant depending on constants $L_U,$ $m_U,$  $M_U$ and $\|f\|_\infty.$
One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$
